---
title: "Assessment 1"
author: "Ruben Adad"
date: "November 15, 2014"
output: pdf_document
---

# Loading and preprocessing the data
```{r loadddata}
setwd("~/Documents/CURSOS/Reproducible Research/RepData_PeerAssessment1")
activity <- read.csv("activity.csv", stringsAsFactors=F)
summary(activity)
head(activity)
```


# What is mean total number of steps taken per day?
The following histogram shows the distribution of the total number of steps taken each day.
The distribution **includes missing values**.
```{r part01a}
library(dplyr)
grp <- group_by(activity, date) 
tot_steps <- summarise(grp, Total_steps = sum(steps))
tot_steps$date <- as.Date(tot_steps$date, "%Y-%m-%d")
hist(tot_steps$Total_steps, main="total number of steps taken each day - includes missing values", xlab="daily steps", col="red")
avg_steps <- summarise(grp, Mean = mean(steps), Median = median(steps))
```
### The mean and median total number of steps taken per day including missing values
```{r part01b, results='asis'}
library(xtable)
print(xtable(as.data.frame(avg_steps), caption="mean and median steps taken per day"), html.table.attributes = list('border="2" bordercolor="blue" style=width:50%'), type="html")
````

# What is the average daily activity pattern?
The following line chart shows the average number of steps taken per 5 minute interval. It also **includes missing values**.
```{r part02}
grp <- group_by(activity, interval)
avg_interval <- summarise(grp, Mean = mean(steps, na.rm=T), Median = median(steps, na.rm=T))
plot(avg_interval$interval, avg_interval$Mean, type="l", lwd=2, col="blue", main="average number of steps taken", xlab="date", ylab="average steps per interval", cex.axis=0.8)
abline(h=c(25,50,75,100,125,150,175,200), lty=3)
max_interval <- subset(avg_interval, avg_interval$Mean == max(na.omit(avg_interval$Mean)), select=c(interval, Mean))
```
The 5-minute interval with the maximum number of steps is: __`r max_interval[1]`__ with mean __`r max_interval[2]`__.

# Imputing missing values
First I check for missing values in all the rows of the _activity_ dataset. 
```{r part03a}
cc <- table(complete.cases(activity))
missing_values <- cc["FALSE"]
````
The total number of rows with missing values is: `r missing_values`  
Then I calculate the mean of all the non-missing values of the _steps_ variable. 
And assign it to all the missing values of de _Mean_ attribute in the *avg_steps* data frame.  
Next I paste the *avg_steps* data frame to the _activity_ data frame using a left join by _date_.  
Finally, I assign the _Mean_ variable from *avg_steps* to the missing values of the _steps_ variable in the _activity_ data frame and drop the _Mean_ and _Median_ columns.
```{r part03b}
avg_steps$Mean[which(complete.cases(avg_steps) %in% "FALSE")] <- mean(na.omit(activity$steps))
new_activity <- left_join(activity, avg_steps)
new_activity$steps[which(complete.cases(new_activity$steps) %in% "FALSE")] <- new_activity$Mean[which(complete.cases(new_activity$steps) %in% "FALSE")]
new_activity <- subset(new_activity, select=c("date", "steps", "interval"))
```
Now I repeat the sum of _steps_ group by date to create the histogram **without missing values**. 
And then to print the mean and median of the number of steps per day.  
**As can be seen this result is different from the previous one which includes missing values.**
```{r part03c}
grp <- group_by(new_activity, date) 
tot_steps <- summarise(grp, Total_steps = sum(steps))
tot_steps$date <- as.Date(tot_steps$date, "%Y-%m-%d")
hist(tot_steps$Total_steps, main="total number of steps taken each day - excluding missing values", xlab="daily steps", col="steelblue")
avg_steps <- summarise(grp, Mean = mean(steps), Median = median(steps))
```
### The mean and median total number of steps taken per day excluding missing values
```{r part03d, results='asis'}
print(xtable(as.data.frame(avg_steps), caption="mean and median steps taken per day"), html.table.attributes = list('border="2" bordercolor="blue" style=width:50%'), type="html")
````

# Are there differences in activity patterns between weekdays and weekends?
I use the function "isWeekend" from the "timeDate" package to determine if a date is weekend or weekday.
Then I replace the logical values TRUE/FALSE with weekend/weekday respectively.  
I generate the comparative graph using ggplot.  
**As can be seen from the red trendline, weekend and weekdays have different patterns.** 
```{r part04}
library(timeDate)
library(ggplot2)
new_activity$weekend <- factor(isWeekend(as.Date(new_activity$date)))
new_activity$weekend <- gsub("TRUE", "weekend", new_activity$weekend)
new_activity$weekend <- gsub("FALSE", "weekday", new_activity$weekend)
grp <- group_by(new_activity, weekend, interval)
avg_interval <- summarise(grp, Mean = mean(steps), Median = median(steps))
g <- ggplot(avg_interval, aes(interval, Mean))
g + geom_line(color="steelblue") + facet_grid(weekend ~ .) + labs(y = "average steps per interval") + geom_smooth(method="lm", color="red")
```

